---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# load package from source when building README
devtools::load_all()
```

# seader

<!-- badges: start -->
<!-- badges: end -->

Versioned municipal data infrastructure for demographic and social analysis in Brazil.

The **seader** package provides functions to access consolidated and versioned public datasets for Brazilian municipalities.
The datasets are published and maintained in the [`seader-data`](https://github.com/fujisima/seader-data) repository.

Currently available:

- 📊 Live births by sex (2019–2025)
- Unit of analysis: Municipality (IBGE code)
- Format: Parquet
- Lazy access via Arrow

## Installation

You can install the development version of seader from [GitHub](https://github.com/) with:

``` r
install.packages("remotes")
remotes::install_github("fujisima/seader")
```

## Example

This example demonstrates how to access birth data:

```{r example}
library(seader)
library(dplyr)

nascidosvivos_sexo <- read_birth_sex(year = 2025)

# 3550308 corresponds to the municipality of São Paulo
query_1 <- nascidosvivos_sexo |> filter(cod_ibge == 3550308) 
query_1 |> collect()

nascidosvivos_idade <- read_birth_mother_age(year = 2025)

# 3550308 corresponds to the municipality of São Paulo
query_2 <- nascidosvivos_idade |> filter(cod_ibge == 3550308) 
query_2 |> collect()

```

